Frequency of exposure mediates the facilitation effect of cognateness in bilingual lexical acquisition

### 2023-02-16 manuscript/manuscript.qmd
Fix Appendix

---
title: "Frequency of exposure mediates the facilitation effect of cognateness in bilingual word acquisition"
editor: source
author:
  - name: Gonzalo Garcia-Castro
    orcid: 0000-0002-8553-4209
    affiliations:
      - Center for Brain and Cognition, Universitat Pompeu Fabra
    corresponding: true
  - name: Daniela S. Ávila-Varela
    affiliations:
      - Center for Brain and Cognition, Universitat Pompeu Fabra
  - name: Ignacio Castillejo
    affiliations:
      - Departamento de Psicología, Universidad Autónoma de Madrid
  - name: Núria Sebastian-Galles
    affiliations:
      - Center for Brain and Cognition, Universitat Pompeu Fabra

@@

abstract: |
    Recent studies have reported larger vocabulary sizes for bilingual toddlers learning languages that shared a greater amount of cognates (e.g., English-Dutch). This effect was only observed in the language to which toddlers were exposed the least, and has been suggested to be the result of a cognateness facilitation effect during word acquisition. The specific underpinnings of tyhis effect, and its relationship with the degree of exposure to the language are unclear. In this study, we present vocabulary data from 436 Catalan-Spanish bilinguals and monolinguals aged 12 to 34. We used Bayesian Item Response Theory to estimate participants' probability of acquisition of 604 words, conditional to their word's cognate status and lexical frequency, and the child's age and degree of exposure to each language. We found converging evidence for an earlier age of acquisition for cognate words, mediated by word frequeny and language exposure. Low-frequency words, and words from the language of least exposure benefitted more strongly by their cognate status than high-frequency words. Our findings support an interactive account of bilingual vocabulary acquisition in which the lexical representations in one language interact with the acquisition and processing of words in the other language.

@@

format:
  pdf:
    pdf-engine: xelatex
    whitespace: small
    linestretch: 1.5

@@

In this study, we provide an account for the cognate facilitation on word acquisition rooted in accumulator models of language acquisition [@hidaka2013computational; @mollica2017how]. We build on the assumption that word acquisition is a function of mainly, but not exclusively, the amount of experience a child has accrued with the word-referent pairing [see @kachergis2022standard for review]. As infants accumulate linguistic input, they also accumulate opportunities to associate familiar word-forms to their corresponding referents. Words with high lexical frequency like *flower* [`r floor(childes[childes$token=="flower",]$freq_million)` times per million words in the CHILDES corpus, @macwhinney2000childes] are more likely to be acquired earlier than words with low lexical frequency such as *headphones* (`r floor(childes[childes$token=="headphones",]$freq_million)` times per million words) [e.g., @goodman2008does]: infants might accumulate learning experiences with *flower* faster than with *headphones*. In the case of bilinguals, the amount of exposures to a word-form an infants receives is mediated by their degree of exposure to their languages. Assuming that bilinguals and monolinguals are exposed to an equivalent amount of linguistic input, a Spanish-English bilingual who is exposed to Spanish 80% of the time and to English 20% of the time might need longer to accumulate the same experience with *flower* and *headphones* than an English monolingual does (who listens 100% of the time to English).

@@

```{r}
#| label: results-values
#| echo: false
loo_diff <- loo_compare(model_loos)[2, 1]
loo_diff_se <- loo_compare(model_loos)[2, 2]

@@

marginal_effects_epreds |> 
    mutate(
        age = rescale_variable(age_std, mean(responses$age), sd(responses$age)),
        exposure = factor(
            exposure_std,
            levels = unique(exposure_std),
            labels = paste0("Exposure: ", c("-1 SD", "mean", "+1 SD")),
            ordered = TRUE
        ),
        lv = factor(
            lv_std,
            levels = unique(lv_std),
            labels = paste0(c(0, 50, 100), "% similarity"),
            ordered = TRUE
        )
    ) |> 
    ggplot(aes(age, .value, colour = lv, fill = lv)) +
    facet_grid(.category~exposure) +
    # stat_summary(fun.data = median_hdi, geom = "ribbon",
    #              alpha = 0.5, colour = NA) +
    annotate(geom = "rect",
             xmin = min(responses$age),
             xmax = max(responses$age),
             ymin = -Inf,
             ymax = Inf,
             fill = "grey90",
             colour = NA) +
    geom_vline(xintercept = mean(responses$age), colour = "grey",
               linewidth = 1) +
    geom_hline(yintercept = 0.5,
               colour = "grey",
               linewidth = 1) +
    geom_line(aes(group = interaction(lv, .draw)),
              linewidth = 0.75, alpha = 0.15) +
    stat_summary(fun = median, geom = "line", linewidth = 1) +
    labs(
        x = "Age (months)",
        y = "Probability of acquisition",
        linetype = "Phonological similarity (Levenshtein)",
        colour = "Phonological similarity (Levenshtein)",
        fill = "Phonological similarity (Levenshtein)",
        subtitle = glue("**Phonological similarity**<br> <span style = 'color:{clrs_reorder[5]};'>0% similarity</span>   <span style = 'color:{clrs_reorder[4]};'>50% similarity</span>   <span style = 'color:{clrs_reorder[1]};'>100% similarity</span>")) +
    scale_colour_manual(values = clrs[c(5, 4, 1)]) +
    scale_fill_manual(values = clrs[c(5, 4, 1)]) +
    scale_y_continuous(
        breaks = seq(0, 1, 0.25),
        limits = c(0, 1),
        labels = percent
    ) +
    scale_x_continuous(breaks = seq(5, 50, 5),
                       limits = c(10, 50)) +
    theme(
        legend.position = "none",
        panel.grid = element_blank(),
        panel.border = element_rect(fill = NA, 
                                    colour = "black", 
                                    linewidth = 0.75),
        plot.subtitle = element_textbox_simple(
            fill = "grey90",
            padding = margin(8, 8, 8, 8),
            margin = margin(0, 0, 10, 0)
        ))
```

@@

# Appendix 1: Model predictors {#sec-appendix-predictors .appendix}

@@

Generally, a pair of words from two different languages that share meaning---translation equivalents---are considered as *cognates* if they share etymological origin (e.g., *table* and *taula*, in English and Catalan). The fact that two words share etymological origin frequently leads to them also being form-similar, as reflected by their overlapping orthographic or phonological form. This makes many cognates perceptually similar. In psycholinguistics, the term *cognateness* is often used to refer to the form-similarity that a pair of translation equivalents share, which has been found to impact how bilinguals process such word-forms (e.g., @costa2000cognate; @spivey1999cross). Whether two form-similar translation equivalents are etymologically related or not is arguably tangential to the question of how cognateness affect language processing. For instance, the translation pair *much* and *mucho*, in English and Spanish have the same meaning, and are considerably similar at the orthographical and phonological level. Nonetheless, they come from different Proto-Indo-European roots [@campbell2007glossary]. This does not keep bilinguals from processing such kind of word-forms word forms differently than other translation equivalents with no form-similarity like *dog* and *perro*, in English and Spanish. In the scope of this experiment, we consider any form-similar translation equivalent as a *cognate* pair, regardless of whether both word-forms share etymological origin.

@@

# Appendix 2: Model details {#sec-appendix-model .appendix}

@@

@@

# Appendix 4: Convergence diagnostics {#sec-appendix-diagnostics .appendix}

## Gelman-Rubin diagnostic

```{r}
#| label: fig-model-rhats
#| fig-cap: "Gelman-Rubin (R-hat) convergence diagnostic of the MCMC chains in the Bayesian models."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 3
#| fig-width: 8
map_df(model_rhats, bind_rows, .id = "model") |> 
    mutate(model = paste0("Model ", str_extract(model, "[0-9]"))) |> 
    pivot_longer(-model, 
                 names_to = ".variable",
                 values_to = ".rhat") |>
    filter(str_detect(.variable, "b_|sd_")) |> 
    drop_na() |> 
    ggplot(aes(.rhat, colour = model, fill = model)) +
    facet_wrap(~model, nrow = 1) +
    annotate(geom = "rect",
             fill = "grey80",
             xmin = 1,
             xmax = 1.01,
             ymin = -Inf,
             ymax = Inf) +
    geom_histogram(bins = 10, colour = "white") +
    geom_rug(alpha = 0.5) +
    labs(x = "Gelman Rubin diagnostic (R-hat)",
         y = "Number of MCMC samples") +
    theme(legend.position = "none")
```

## Effective sample size

```{r}
#| label: fig-model-ess
#| fig-cap: "Ratio of effective sample size in the MCMC chains in the Bayesian models."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 3
#| fig-width: 8
map_df(model_neffs, bind_rows, .id = "model") |> 
    mutate(model = paste0("Model ", str_extract(model, "[0-9]"))) |> 
    pivot_longer(-model, 
                 names_to = ".variable",
                 values_to = ".rhat") |>
    filter(str_detect(.variable, "b_|sd_")) |> 
    drop_na() |> 
    ggplot(aes(.rhat, colour = model, fill = model)) +
    facet_wrap(~model, nrow = 1) +
    geom_vline(xintercept = 0.1, colour = "grey", linewidth = 0.75) +
    geom_histogram(bins = 10, colour = "white") +
    geom_rug(alpha = 0.5) +
    labs(x = "Gelman Rubin diagnostic (R-hat)",
         y = "Number of MCMC samples") +
    scale_x_continuous(labels = percent) +
    theme(legend.position = "none")
```

# Appendix 5: Frequency and language exposure as separate predictors {#sec-appendix-doe  .appendix}

>>>>

---
title: "Frequency of exposure mediates the facilitation effect of cognateness in bilingual lexical acquisition"
editor: source
author:
  - name: Gonzalo Garcia-Castro
    orcid: 0000-0002-8553-4209
    affiliations:
      - Center for Brain and Cognition, Universitat Pompeu Fabra
    corresponding: true
  - name: Daniela S. Ávila-Varela
    affiliations:
      - Center for Brain and Cognition, Universitat Pompeu Fabra
  - name: Ignacio Castillejo
    affiliations:
      - Departamento de Psicología, Universidad Autónoma de Madrid
  - name: Núria Sebastian-Galles
    affiliations:
      - Center for Brain and Cognition, Universitat Pompeu Fabra

@@

abstract: |
    One of the most prominent features of the bilingual lexicon is the availability of word-form representations from one language to those from the other language. One instance of such phenomenon is that fact that the form overlap between a word-form and its translation (i.e., cognateness) impacts how bilinguals process it. Recent studies have suggested that cognateness facilitates vocabulary acquisition in bilingual toddlers, who show larger vocabulary sizes when their languages share many cognates, and who also acquire cognates earlier in age than non-cognates. The specific mechanisms underpinning such facilitation effect are unclear. In this study, we present a model of bilingual early lexical acquisition in which cognateness interacts with lexical frequency and language exposure to facilitate the acquisition of low-frequency words. We tested this model against vocabulary data from 436 Catalan-Spanish bilinguals aged 12 to 34 months. We used Bayesian Item Response Theory to estimate participants’ probability of acquisition of 604 words, conditional to the cognate status and lexical frequency of the word-form, and the age and degree of exposure to each language of the toddler. We found converging evidence for an earlier age of acquisition for cognate words, and for such effect being mediated by lexical frequency and language exposure. Low-frequency words, and words from the language of least exposure benefited more strongly by their cognate status than high-frequency words. Overall, our findings support an interactive account of bilingual vocabulary acquisition in which the lexical representations in one language interact with the acquisition and processing of words in the other language.

@@

format:
  pdf:
    pdf-engine: xelatex
    whitespace: small
    linestretch: 1.5
  docx: 
    reference-doc: "../assets/templates/template.docx"

@@

In this study, we provide an account for the cognate facilitation on word acquisition rooted in accumulator models of language acquisition [@hidaka2013computational; @mollica2017how]. We build on the assumption that word acquisition is a function of mainly---but not exclusively---the amount of experience a child has accrued with the corresponding word-referent pairing [see @kachergis2022standard for review]. As infants accumulate linguistic input, they also accumulate opportunities to associate familiar word-forms to their corresponding referents. Words with high lexical frequency like 'flower' [`r format(floor(childes[childes$token=="flower",]$freq_million), big.mark = ",")` times per million words in the CHILDES corpus, @macwhinney2000childes] are more likely to be acquired earlier than words with low lexical frequency such as 'headphones' (`r format(floor(childes[childes$token=="headphones",]$freq_million), big.mark = ",")` times per million words) [e.g., @goodman2008does]. Infants might accumulate learning experiences with 'flower' faster than with 'headphones'. In the case of bilinguals, the amount of exposures to a word-form an infants receives is mediated by their degree of exposure to their languages. Assuming that bilinguals and monolinguals are exposed to an equivalent amount of linguistic input, a Spanish-English bilingual who is exposed to Spanish 80% of the time and to English 20% of the time might need longer to accumulate the same experience with 'flower' and 'headphones' than an English monolingual does (who listens 100% of the time to English).

@@

```{r}
#| label: results-values
#| echo: false

rhats_df <- map_df(model_rhats, bind_rows, .id = "model") |> 
    mutate(model = paste0("Model ", str_extract(model, "[0-9]"))) |> 
    pivot_longer(-model, 
                 names_to = ".variable",
                 values_to = ".rhat") |>
    filter(str_detect(.variable, "b_|sd_")) |> 
    drop_na()

loo_diff <- loo_compare(model_loos)[2, 1]
loo_diff_se <- loo_compare(model_loos)[2, 2]

@@

marginal_effects_epreds |> 
    mutate(
        age = rescale_variable(age_std, mean(responses$age), sd(responses$age)),
        exposure = factor(
            exposure_std,
            levels = unique(exposure_std),
            labels = paste0("Exposure: ", c("-1 SD", "mean", "+1 SD")),
            ordered = TRUE
        ),
        lv = factor(
            lv_std,
            levels = unique(lv_std),
            labels = paste0(c(0, 50, 100), "% similarity"),
            ordered = TRUE
        )
    ) |> 
    ggplot(aes(age, .value, colour = lv, fill = lv)) +
    facet_grid(.category~exposure) +
    # stat_summary(fun.data = median_hdi, geom = "ribbon",
    #              alpha = 0.5, colour = NA) +
    annotate(geom = "rect",
             xmin = min(responses$age),
             xmax = max(responses$age),
             ymin = -Inf,
             ymax = Inf,
             fill = "grey90",
             colour = NA) +
    geom_vline(xintercept = mean(responses$age), colour = "grey",
               linewidth = 1) +
    geom_hline(yintercept = 0.5,
               colour = "grey",
               linewidth = 1) +
    geom_line(aes(group = interaction(lv, .draw)),
              linewidth = 0.75, alpha = 0.15) +
    stat_summary(fun = median, geom = "line", linewidth = 1) +
    labs(
        x = "Age (months)",
        y = "Probability of acquisition",
        linetype = "Phonological similarity (Levenshtein)",
        colour = "Phonological similarity (Levenshtein)",
        fill = "Phonological similarity (Levenshtein)",
        subtitle = glue("**Phonological similarity**<br> <span style = 'color:{clrs_reorder[5]};'>0% similarity</span>   <span style = 'color:{clrs_reorder[4]};'>50% similarity</span>   <span style = 'color:{clrs_reorder[3]};'>100% similarity</span>")) +
    scale_colour_manual(values = clrs_reorder[c(5, 4, 3)]) +
    scale_fill_manual(values = clrs_reorder[c(5, 4, 3)]) +
    scale_y_continuous(
        breaks = seq(0, 1, 0.25),
        limits = c(0, 1),
        labels = percent
    ) +
    scale_x_continuous(breaks = seq(5, 50, 5),
                       limits = c(10, 50)) +
    theme(
        legend.position = "none",
        panel.grid = element_blank(),
        panel.border = element_rect(fill = NA, 
                                    colour = "black", 
                                    linewidth = 0.75),
        plot.subtitle = element_textbox_simple(
            fill = "grey90",
            padding = margin(8, 8, 8, 8),
            margin = margin(0, 0, 10, 0)
        ))
```

@@

{{< pagebreak >}}

# Appendix 1: model predictors {#sec-appendix-predictors .appendix}

@@

Generally, a pair of words from two different languages that share meaning---translation equivalents---are considered as *cognates* if they share etymological origin (e.g., *table* and *taula*, in English and Catalan). The fact that two words share etymological origin frequently leads to them also being form-similar, as reflected by their overlapping orthographic or phonological form. This makes many cognates perceptually similar. In psycholinguistics, the term *cognateness* is often used to refer to the form-similarity that a pair of translation equivalents share, which has been found to impact how bilinguals process such word-forms (e.g., @costa2000cognate; @spivey1999cross). Whether two form-similar translation equivalents are etymologically related or not is arguably tangential to the question of how cognateness affect language processing. For instance, the translation pair *much* and *mucho*, in English and Spanish have the same meaning, and are considerably similar at the orthographical and phonological level. Nonetheless, they come from different Proto-Indo-European roots [@campbell2007glossary]. This does not keep bilinguals from processing such kind of word-forms word forms differently than other translation equivalents with no form-similarity like 'dog' and *perro*, in English and Spanish. In the scope of this experiment, we consider any form-similar translation equivalent as a *cognate* pair, regardless of whether both word-forms share etymological origin.

@@

{{< pagebreak >}}

# Appendix 2: model details {#sec-appendix-model .appendix}

@@

{{< pagebreak >}}

@@

{{< pagebreak >}}

# Appendix 4: convergence diagnostics {.appendix}

We used Stan [@carpenter2017stan] as the probabilistic language behind the estimation of our Bayesian models in this study, with `brms` as its R interface [@burkner2017brms]. This language implements the Markov Chain Monte-Carlo (MCMC) algorithm to explore the posterior distribution of the model. Broadly, this algorithm is used to iteratively sample the joint sampling space of the parameters to be estimated in the model, and compute, for each value sampled, its likelihood under some probability distribution previously defined. We run `r dim(model_fit_4$fit)[2]` MCMC chains, each `r dim(model_fit_4$fit)[1]` iterations long each. The correct performance of this algorithm is critical to the quality of the statistical evidence to which the outcomes of the model lead.

One way to diagnose the behaviour of MCMC is to inspect whether the different MCMC chains (if more than one) have converged to a similar region of the posterior. The Gelman-Rubin diagnostic [$\hat{R}$ or R-hat @gelman1992inference] provides a measure of chain convergence by comparing the variance within each chain *versus* the variance between each chain. Both are expected to be identical when chains have perfectly converged, so that $\hat{R} = 1$. Values lower than 1.01 are recommended, while values higher than 1.05 indicate that chains might have trouble converging and therefore the estimated parameters must be taken with caution. @fig-model-rhats-neffs (A) shows the distribution of $\hat{R}$ values for the coefficients of the fixed effect of our models, which we used for statistical inference. Most values are lower than 1.01, and never higher than 1.05, which provides evidence of successful MCMC convergence.

Another diagnostic of good MCMC converge is the ratio of effective sample size to total sample size ($N_{eff}/N$), which indicates the proportion of samples in the chain that resulted from a non-divergent transition. Values closer to 1 are ideal, as they indicate that all posterior samples from the MCMC were used to estimate the posterior distribution of the parameter. Values larger than 0.1 are recommended. @fig-model-rhats-neffs (B) shows the distribution of the effective sample sizes of the coefficients of the fixed effects in our models. Most values are larger than 0.1, although model 0 ($\mathcal{M}_0$) accumulates most effective sample sizes close to 0.1.

```{r}
#| label: fig-model-rhats-neffs
#| fig-cap: "Gelman-Rubin (R-hat) convergence diagnostic of the MCMC chains in the Bayesian models."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 9
map_df(model_rhats, bind_rows, .id = "model") |> 
    mutate(model = paste0("Model ", str_extract(model, "[0-9]"))) |> 
    pivot_longer(-model, 
                 names_to = ".variable",
                 values_to = ".rhat") |>
    filter(str_detect(.variable, "b_|sd_")) |> 
    drop_na() |> 
    ggplot(aes(.rhat, colour = model, fill = model)) +
    facet_wrap(~model, nrow = 1) +
    geom_vline(xintercept = 1.01, colour = "grey", linewidth = 0.75) +
    geom_histogram(bins = 12, colour = "white") +
    geom_rug(alpha = 0.5) +
    labs(x = "Gelman Rubin diagnostic (R-hat)",
         y = "MCMC samples",
         title = "Gelman-Rubin (R-hat) diagnostic") +

    map_df(model_neffs, bind_rows, .id = "model") |> 
    mutate(model = paste0("Model ", str_extract(model, "[0-9]"))) |> 
    pivot_longer(-model, 
                 names_to = ".variable",
                 values_to = ".rhat") |>
    filter(str_detect(.variable, "b_|sd_")) |> 
    drop_na() |> 
    ggplot(aes(.rhat, colour = model, fill = model)) +
    facet_wrap(~model, nrow = 1) +
    geom_vline(xintercept = 0.1, colour = "grey", linewidth = 0.75) +
    geom_histogram(bins = 10, colour = "white") +
    geom_rug(alpha = 0.5) +
    labs(x = "Gelman Rubin diagnostic (R-hat)",
         y = "MCMC samples",
          title = "Ratio of effective sample size") +
    scale_x_continuous(labels = percent) +

    plot_layout(ncol = 1) &
    plot_annotation(tag_levels = "A") &
    theme(legend.position = "none",
          panel.grid = element_blank())
```

# Appendix 5: frequency and language exposure as separate predictors {.appendix}
